---
title:  'Week 5 - R Vector/matrix manipulations and math, speed considerations'
author: "Robert Settlage"
date: "2021-10-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
header-includes: \setlength\parindent{24pt} \usepackage{MnSymbol} \usepackage{mathrsfs}
---


```{r setup, eval=T, echo=F, include=F}
  library(ggplot2); library(microbenchmark)
```

## Today's Agenda

- Clean up Git and Graphics 
- R functions  
- matrix operations 


### Now lets add a branch.

```
git branch mynewstuff  
## what branch are we on?  
git branch --list  
echo "\nadd a new line or two\n" >>README.md
cat README.md
git add README.md
git commit -m "adding new lines"  
## why did I need to do the last?  
git switch mynewstuff  
cat README.md
```

What happened?  How do we fix it?

### git pull

```
git pull
git pull origin main
cat README.md
```

The reason this is cool is you can have several versions of a project active and available at the same time.  In the same place, and the exact version is changable via a `git switch`


## R and useful matrix stuff

1. matrix mult %*%
2. transpose t()
3. create diagonal matrix diag()
    + this one has two cases
4. matrix inverse solve()
5. determinate det()
6. rank matA <- qr(A); matA$rank
7. t(x) %*% y crossprod()

More comprehensive list here:
<https://www.statmethods.net/advstats/matrix.html>


## vectors and matrices in R

Super easy to create:

```{r echo=T, eval=F, include=T}
    a <- c(1:5)
    B <- matrix(1:10,ncol=2)
    a
    B
```

## basic operations


```{r echo=T, eval=F, include=T}

## add scalar
a
a + 5
## multiply by scalar
B
B * 5
## multiply vectors/matrix
a * a
a * B
B * a
## length of vector
sqrt(sum(a*a))
```

## matrix math

```{r echo=T, eval=F, include=T}
## transpose
t(a)
t(t(B))
## matrix multiply
a %*% B
B %*% a
t(B) %*% a
## diag
diag(1,3)
diag(c(1,2,3))
diag(B)
## solve
C <- matrix(1:9 + rnorm(9,0,1e-3),ncol=3) 
C %*% solve(C)
## trace
sum(diag(C))
## det
det(C)
## Kronecker product
I <- diag(3)
J <- matrix(1,3,3)
blockDiag <- kronecker(I, J)
```

## Using duality to do stuff

colMeans.  What if we want the means of a matrix by columns.  Of course there are functions for this, but, for kicks, can we do this with some mixed matrix art ... ??

```{r eval=F, echo=T, include=T}

C <- matrix(1:20 + rnorm(1600,0,1e-3),ncol=40)
colMeans(C) ## presumably optimized version
ones <- rep(1,40)
t(ones) %*% C / 40 ## manual but using matrix operations

library(microbenchmark)

#for kicks
microbenchmark(result1<-{colMeans(C)},
               result2<-{t(ones) %*% C / 40},times = 100, unit = "ms")

```

## Making things faster:

A lot of effort has gone into making matrix math faster on computers.  Again, this is a guided tour, so I will show you a few things and hope this sparks your curiousity enough to hit google:

```{r echo=T, eval=F, include=T}

    A = matrix(rnorm(20*40000, mean=0, sd=5), 20, 40000)
    B = matrix(rnorm(20*40000, mean=0, sd=5), 20, 40000)
    time1 <- system.time({t(A)%*%B})
    time2 <- system.time({crossprod(A,B)})
    print(rbind(time1,time2))

```

Don't invert matrices:

<https://www.johndcook.com/blog/2010/01/19/dont-invert-that-matrix/>
<https://www.r-bloggers.com/dont-invert-that-matrix-why-and-how/>

```{r echo=T, eval=F, include=T}

    n <- 5000
    A = matrix(rnorm(n*n, mean=0, sd=5), n, n)
    x <- rnorm(n)
    system.time({b_inverting <- solve(A)%*%x})
    system.time({b_solving <- solve(A,x)})

    max(abs(b_inverting - b_solving))
```

## For those interested in boosting their local R

Not all installations of R come accelerated:  

*There are several highly optimized libraries that can be used instead of the default base libraries. These libraries are optimized to take advantage of the hardware they are run on, and can be significatantly faster than the base implementation (operations such as Matrix multiplications may be over 40 times faster*  
<https://csantill.github.io/RPerformanceWBLAS/>

## What about in Python??

Should point on the cool cheat sheet on this:

<https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwiJjbnJ7rX6AhWsFlkFHchNDQwQFnoECAcQAQ&url=https%3A%2F%2Fraw.githubusercontent.com%2Frstudio%2Fcheatsheets%2Fmain%2Freticulate.pdf&usg=AOvVaw3zphxsBi9OZLtBiNaUO6I7>

In Python, base python doesnt have great support for matrices, dataframes etc, so you need to use a package, generally numpy or pandas.

```{r python_setup, eval=TRUE, echo=TRUE}

library(reticulate)
conda_list()
use_condaenv("r-reticulate")

```

## vectors and matrices in Python

Super easy to create:

```{python echo=T, eval=F, include=T}
   import numpy as np
   # vectors are 1-D arrays
   a = np.array([1,2,3,4])
   B = np.array([[1,2], [3,4]])
   
   print("a:\n", a)
   print("B:\n", B)
   
   ## dim in python is shape
   a.shape
   B.shape
   ## check out reshape
   B.reshape((4,))
   print("B:\n", B)
   np.arange(10).reshape((2,5))
   oneCol = np.ones((5,))  # a single vector of ones
   zeroCols = np.zeros((5,2))  # two columns of zeros
   np.column_stack((oneCol, zeroCols))  # 5x3 columns
```

## basic operations


```{python echo=T, eval=F, include=T}

## add scalar
a
a + 5
## multiply by scalar
B
B * 5
## multiply vectors/matrix
a * a
a * B
B * a
## note element positions are zero based
a[2]
B[:,1]
```

## matrix math

```{python echo=T, eval=F, include=T}
## transpose
a.T
B.T
## matrix multiply
B = np.arange(16).reshape((4,4))
a @ B
B @ a
np.matmul(a,B)
## diag
np.diag([1,2,3,4])
np.diag(B)
## solve
B = np.arange(4).reshape(2,2)
np.linalg.inv(B)
## trace
B.trace()
## det
np.linalg.det(B)
## Kronecker product
I = np.identity(3)
blockDiag = np.kron(I,B)
```

## Panda = all things data frame

Pandas derives from panel data, but its much more.  It is great for data frames, series, panels

```{python eval=FALSE}
import pandas as pd
data = [['Alex',10],['Bob',12],['Clarke',13]]
df = pd.DataFrame(data,columns=['Name','Age'])
df
df['Name']
df.iloc[:,1]
type(df)
#dir(pd)
dir(df)

from palmerpenguins import load_penguins
penguins = load_penguins()
type(penguins)
penguins.describe
penguins.dtypes
penguins.shape
penguins.iloc[:,2:5]
penguins.iloc[:,[2,6]]
```
